---
title: "Mime-16s-emp"
author: "Clara Jégousse"
date: "3/3/2021"
output:
  pdf_document:
    toc: yes
  word_document:
    toc: yes
  html_document:
    toc: yes
---

# Load libraries



```{r, message=FALSE, warning=FALSE}
library(devtools)
library(ggplot2)
library(ggpubr)
library(dada2)
library(phyloseq)
library(reshape2) # to use melt
library(phylosmith)
```

# Visual setting

Import variables and functions for consistent plots.

```{r}
source_url("https://raw.githubusercontent.com/clarajegousse/mime-16s/main/scripts/visual-settings.r")
```

# Load data 

The results of dada2 sequence processing were organized into a phyloseq object containing all 1397 samples amplified with the EMP primers with metadata from Hafro. 

```{r, echo=TRUE}
ps <- readRDS("/Users/Clara/Projects/mime-16s/global-ps-emp.rds")

dna <- Biostrings::DNAStringSet(taxa_names(ps))
names(dna) <- taxa_names(ps)
ps <- merge_phyloseq(ps, dna)
taxa_names(ps) <- paste0("ASV", seq(ntaxa(ps)))
ps

```

All metadata including measures from Hafro.
```{r, warning=FALSE}
sample_variables(ps)
```

```{r}
rank_names(ps)
```

Number of taxa
```{r, warning=FALSE}
ntaxa(ps)
```

# Deal with missing values in metadata with imputation of the mean of the two surrounding values

```{r, message=FALSE, warning=FALSE}
library(imputeTS)
sample_data(ps)$po4.avg <- round(na_ma(sample_data(ps)$po4.avg, k = 1), digits = 2)

sample_data(ps)$sio2.avg <- round(na_ma(sample_data(ps)$sio2.avg, k = 1), digits = 2)

sample_data(ps)$no3.avg <- round(na_ma(sample_data(ps)$no3.avg, k = 1), digits = 2)

sample_data(ps)[is.na(sample_data(ps)$chl.a),]$chl.a <- 0
sample_data(ps)[is.na(sample_data(ps)$phaeo),]$phaeo <- 0
sample_data(ps)[is.na(sample_data(ps)$rfsu),]$rfsu <- 0
```

# Filtering

Filtering prevents spending time analyzing unreliable data, background noise (taxa that are actually just artifacts of the data collection process) and taxa that are seen rarely among samples.

## Filtering samples

The objective is to remove samples with very low numbers of reads. So first we investigate the overall sequencing depths per sample setting a arbitrary threshold of 2000 reads as the minimum numbers of reads for a sample. 

```{r, message=FALSE, warning=FALSE, fig.fullwidth=TRUE}
# number of reads per samples
reads <- as.data.frame(sample_sums(ps))
colnames(reads) <- c("total")
reads$run <- sample_data(ps)$run
reads$sample <- rownames(reads)

gghistogram(reads, x = "total",
            add = "mean", rug = TRUE,
            bins = 50,
            color = MediumGrey, fill = MediumGrey,
            palette = Palette1) +
  geom_vline(xintercept = 2000, linetype = 2, col = DarkGrey) +
  clean_theme + theme(axis.text.x = element_text(angle = 0, vjust = 0, hjust=.5))+
  xlab("Total number of reads") + ylab("Samples") +
  labs(title = "Distribution of reads", 
       subtitle = "All samples from all 11 MiSeq runs", 
       caption = paste0("MiSeq runs (n = ", length(unique(reads$run)), ")\n",
                        "Samples (n = ", dim(reads)[1], ")"))
```

We investigate the sequencing depth per samples for each MiSeq run.

```{r, fig.fullwidth=TRUE}
gghistogram(reads, x = "total",
            add = "mean", rug = TRUE,
            color = "run", fill = "run",
            bins = 50,
            #color = MediumGrey, fill = MediumGrey,
            palette = Palette1[-c(11,12)]) + 
  facet_wrap(~run) +
  geom_vline(xintercept = 2000, linetype = 2, col = DarkGrey) +
  clean_theme + theme(legend.position = "none") +
  xlab("Reads") + ylab("Samples") +
  labs(title = "Distribution of reads", 
       subtitle = "Samples amplified with EMP primers for each MiSeq run", 
       caption = paste0("MiSeq runs (n = ", length(unique(reads$run)), ")\n",
                        "Samples (n = ", dim(reads)[1], ")"))
```

The plots confirm that we can filter out samples containing less than 2000 reads. 

```{r}
# based on the plots above define the minimum number of reads per sample
min.reads <- 2000
smp.keeper <- reads[reads$total >= min.reads,]$sample

ps0 <- ps %>%
  subset_samples(rownames(sample_data(ps)) %in% smp.keeper)
ps0 
```

The total number of samples removed because they contained less than 2000 reads.

```{r}
length(reads[reads$total <= min.reads,]$sample)
```


## Filter taxa

The samples were amplified with the EMP primers therefore it is reasonable to filter taxonomic features for which a high-rank taxonomy could not be assigned - like "Uncharacterized" at the Kingdom level. Such ambiguous features in this setting are almost always sequence artifacts that do not exist in nature. Here we remove everything that was not characterised as "Bacteria" at the kingdom level.

```{r}
# because these were assigned with Silva
ps0 <- subset_taxa(ps0, Kingdom %in% c("Bacteria"))

# check the phyla within Bacteria
table(tax_table(ps0)[, "Phylum"], exclude = NULL)
```

## Prevalence filtering

Prevalence filtering is unsupervised, relying only on the data in this experiment, and a parameter that we choose after exploring the data. Thus, this filtering step can be applied even in settings where taxonomic annotation is unavailable or unreliable.

First, explore the relationship of prevalence and total read count for each feature. Sometimes this reveals outliers that should probably be removed, and also provides insight into the ranges of either feature that might be useful.

Define prevalence of each taxa (in how many samples did each taxa appear at least once).

```{r, message=FALSE ,warning=FALSE, fig.fullwidth=TRUE}
# Define prevalence of each taxa
# (in how many samples did each taxa appear at least once)
prev0 = apply(X = otu_table(ps0),
              MARGIN = ifelse(taxa_are_rows(ps0), yes = 1, no = 2),
              FUN = function(x){sum(x > 0)})
prevdf = data.frame(Prevalence = prev0,
                    TotalAbundance = taxa_sums(ps0),
                    tax_table(ps0))

# Define prevalence threshold as 1% of total samples
prevalenceThreshold = round(0.01 * nsamples(ps0), digits = 0)
prevalenceThreshold
```

```{r}
# Execute prevalence filter, using `prune_taxa()` function
ps1 = prune_taxa((prev0 > prevalenceThreshold), ps0)
ps1
```

```{r}
table(prevdf$Phylum)
keepPhyla = table(prevdf$Phylum)[(table(prevdf$Phylum) > 5)]
prevdf1 = subset(prevdf, Phylum %in% names(keepPhyla))

# Filter entries with unidentified Phylum.
ps2 = subset_taxa(ps1, Phylum %in% names(keepPhyla))
ps2
```

```{r}
ggplot(prevdf1, aes(TotalAbundance, Prevalence / nsamples(ps0), color=Phylum)) +
  geom_hline(yintercept = 0.05, alpha = 0.5, linetype = 2) + 
  geom_point(size = 2, alpha = 0.7) +
  scale_x_log10() +  
  tax_color_scale(ps0, "Phylum") +
  xlab("Total Abundance") + ylab("Prevalence [Frac. Samples]") +
  facet_wrap(~Phylum) + clean_theme + theme(legend.position="none") 
```

## Agglomerate taxa at the Genus level

There is a lot of species, sub-species, or strains with functional redundancy in the marine microbial community, so we can agglomerate the data features corresponding to closely related taxa (here at the Genus level) as we looking at overall patterns. 

```{r}
taxGlomRank = "Genus"
length(get_taxa_unique(ps2, taxonomic.rank = taxGlomRank))

ps3 = tax_glom(ps2, taxrank = taxGlomRank)
```

## Filter specific samples

For now, let's focus on one survey.

```{r, warning=FALSE}
# ps4 <- subset_samples(ps3, is.na(cruise) == FALSE & cruise != "B8-2010")
ps4 <- subset_samples(ps3, cruise == "B7-2017")
ps4
```

# Data normalisation

## Normalising OTU abundance 

Normalize number of reads in each sample using median sequencing depth (cf. Daniel Vaulot).

```{r, warning=FALSE, fig.fullwidth=TRUE}
# with microbiomeSeq
# ps4n <- normalise_data(ps4, norm.method = "relative")
# phylosmith
# ps4n <- relative_abundance(ps4)

total = median(sample_sums(ps4))
standf = function(x, t=total) round(t * (x / sum(x)))
ps4n = transform_sample_counts(ps4, standf)
ps4n
```

# Beta diversity

```{r, warning=FALSE, fig.fullwidth=TRUE, fig.width=10}
plot_bar2(ps4n, x = "Sample", y = "Abundance", fill = "Phylum")
```

```{r, warning=FALSE, fig.fullwidth=TRUE, fig.width=10}
plot_bar2(ps4n, x = "Sample", y = "Abundance", fill = "Phylum") + facet_grid(zone~.)
```

# Alpha diversity

TODO: ANOVA to confirm is alpha diversity is significantly different between zones

## Photic zone

```{r, warning=FALSE, fig.fullwidth=TRUE}
p <- plot_richness(ps4n, measures = c("Chao1", "Shannon", "ACE"), 
              x = "zone", color = "zone") +
  scale_color_manual(values = c(Grass, Mint, Aqua, Jeans))

p + geom_boxplot(data = p$data, aes(x = zone, y = value), color = DarkGrey, 
    alpha = 0.5) + clean_theme
```

## Region

```{r, warning=FALSE, fig.fullwidth=TRUE}
p <- plot_richness(ps4n, measures = c("Chao1", "Shannon", "ACE"), 
              x = "region", color = "region") +
  scale_color_manual(values = c(Grapefruit, Jeans))

p + geom_boxplot(data = p$data, aes(x = region, y = value), color = DarkGrey, 
    alpha = 0.5) + clean_theme
```

# Ordination

```{r, message=FALSE, results='hide'}
ps4n.ord <- ordinate(ps4n, "NMDS", "bray")
```


## Display ASVs

```{r, warning=FALSE, fig.fullwidth=TRUE}
 plot_ordination(ps4n, ps4n.ord, type = "taxa", 
                 color = "Class", title = "AVSs") +
  tax_color_scale(ps4, "Class") +
  clean_theme + theme(aspect.ratio = 1,
                      axis.text.x = element_text(angle = 0, 
                                                 hjust = .5, 
                                                 vjust = 0),
                      legend.position = "right")
```

## Display samples

```{r, warning=FALSE, fig.fullwidth=TRUE}
plot_ordination(ps4n, ps4n.ord, type="samples", color="zone", 
                  shape="region", title="Samples") + geom_point(size=3) +
  scale_color_manual(values = c(Grass, Mint, Aqua, Jeans)) +
  clean_theme + theme(aspect.ratio = 1,
                      axis.text.x = element_text(angle = 0, 
                                                 hjust = .5, 
                                                 vjust = 0),
                      legend.position = "right")
```


## AVSs and samples

```{r, warning=FALSE, fig.fullwidth=TRUE}
  plot_ordination(ps4n, ps4n.ord, type="split", color = "Class", 
                  shape = "region", title="biplot", label = "stn") +  
  geom_point(size=3) + 
  tax_color_scale(ps4, "Class") +
  clean_theme + theme(aspect.ratio = 1,
                      axis.text.x = element_text(angle = 0, 
                                                 hjust = .5, 
                                                 vjust = 0),
                      legend.position = "right")
```


# Correlation matrix

```{r, warning=FALSE, fig.fullwidth=TRUE, fig.height=10, message=FALSE, echo=FALSE}
variable_correlation_heatmap(ps4n, variables = c('temp.avg', 'salt.avg', 'po4.avg', 'sio2.avg', 'no3.avg'),
treatment = c('region'),
classification = 'Phylum', method = 'spearman', cores = 4,
significance_color = 'black') + clean_theme
```

